---
title: "Activity 1"
author: "Brooke Wolford"
date: "2025-12-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We are reading in data from the National Health and Nutrition Examination Survey from August 2021-August 2023. The complete set of data is available [here](https://wwwn.cdc.gov/nchs/nhanes/search/datapage.aspx?Component=Questionnaire&Cycle=2021-2023).

These are in a x.pt format which is a SAS Transport file format for data sharing and can be read into R with the `haven` package.

The documentation for the files can be found here:
[diabetes](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DIQ_L.htm)
[physical activity](https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/PAQ_L.htm)

The join functions are important within dplyr. You can read more [here](https://dplyr.tidyverse.org/reference/mutate-joins.html).

```{r}

#install.packages("haven")
library(haven)

#diabetes
dia<-read_xpt("data/DIQ_L.xpt")

#physical activity
pa<-read_xpt("data/PAQ_L.xpt")

#diabetes data has more samples than physical activity, so let's do a left join with physicial actvitiy as the anchoring file
df<-left_join(pa,dia)
